The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-08-05 paper/iphas-dr2-rev1.tex
Tinkered with figure placement

An external validation of the astrometry 
has been carried out by comparing our positions against the
United States Naval Observatory CCD Astrograph Catalog 
\citep[UCAC4;][]{UCAC4}.
Fig.~\ref{fig:astrometry} shows the distribution
of the position offsets against UCAC4
computed for 1.3 million stars
in the magnitude range $13 < r < 15$,
which is the range where both surveys overlap
and where the formal mean error of UCAC4
is better than 50~mas \citep{UCAC3}.
We find the mean position offset to be $0\farcs 09\pm0\farcs 07$,
which is satisfactory for our purposes.
%We note that both UCAC4 and 2MASS -- and hence also IPHAS --
%not 100% true due to bulk pm of calibrators - Mike
%are calibrated in the International Celestial Reference System (ICRS).
%We note that the mean residual is  
%is $-0.006 \pm 0.086$ in R.A.
%and $0.012\pm0.073$ in Dec.

@@

The entire data set -- comprising 2.5~terabyte of FITS files --
was then transferred to the University of Hertfordshire
for the purpose of transforming the raw
detection tables into a proper source catalogue which
(i) is quality-controlled,
(ii) is homogeneously calibrated, and 
(iii) contains user-friendly columns and warning flags.
It is these post-processing steps which distinguish this release
from the IDR, which (i) enforced less stringent quality limits,
(ii) did not offer a global calibration,
and (iii) provided a catalogue in which duplicate detections
of unique sources were not flagged.
These improvements are explained next.}

@@

@@

\begin{table*}
\vspace{2.5cm}
\caption{\bf Recommended quality criteria 
for selecting objects with accurate colours.
These criteria serve to identify objects detected 
at S/N$\,>\,$10 in all three bands without being saturated,
and with the added requirement that the photometric measurements
need to be consistent across different aperture sizes.
The 86~million objects which satisfy these criteria
have been flagged in the catalogue using the column named \emph{a10}
(short for ``all-band 10$\sigma$'').}
\begin{tabular}{p{8cm}lp{6cm}}
  \hline
  Quality criterion & Rows passed & Description \\
  \hline
  $rErr\,<\,0.1$ {\sc and} $iErr\,<\,0.1$ {\sc and} $haErr\,<\,0.1$ &
  109 million (50\%) &
  Require the photon noise to be less than 
  0.1 mag in all bands (i.e. S/N$>$10).
  This implicitly requires a detection in all three bands.  \\
  {\sc not} \emph{saturated} &
  158 million (72\%) &
  The brightness must not exceed the nominal saturation limits.
  \\
  $|r- rAperMag1| 
  < 3\sqrt{rErr^2 + rAperMag1Err^2} 
  + 0.03$ &
  176 million (80\%) &
  Require the $r$ magnitude measured 
  in the default 2\farcs 3-diameter aperture
  to be consistent with the measurement 
  made in the smaller 1\farcs 2 aperture,
  albeit tolerating a 0.03 mag systematic error.
  This will reject sources for which the background
  subtraction or the deblending procedure
  was not performed reliably. \\
  $|i- iAperMag1| 
  < 3\sqrt{iErr^2 + iAperMag1Err^2} 
  + 0.03$ &
  183 million (84\%) &
  Same as above for $i$. \\
  $|ha- haAperMag1| < 
  3\sqrt{haErr^2 + haAperMag1Err^2} 
  + 0.03$ &
  158 million (72\%) &
  Same as above for \ha. \\
  \hline
  All of the above (flagged as {\bf\emph{a10}}) &
  86 million (39\%) & \\
  \hline
\end{tabular}
\label{tab:a10}

\vspace{2.5cm}
\caption{\bf Additional quality criteria which are recommended
for applications which require objects to be
single, unconfused point sources with accurate colours.
The 59 million sources which satisfy these criteria 
have been flagged using the column named \emph{a10point}.}
\begin{tabular}{p{8cm}lp{6cm}}
  \hline
  Quality criterion & Rows passed & Description \\
  \hline
   \emph{a10} &
   86 million (39\%) &
   The object must satisfy the criteria for accurate colours listed in Table~\ref{tab:a10}. \\

@@

%In producing a point source catalogue, spurious detections 
%will inevitably be made. A fraction of these will mimic stellar 
%sources closely enough that they cannot be excluded using the 
%recorded parameters.

@@

To appraise the level at which our photometry is affected by such systematics,
we can exploit the secondary measurements which are present in the catalogue
(i.e. \emph{r} vs \emph{r2}, \emph{i} vs \emph{i2}, \emph{ha} vs \emph{ha2}).
In Fig.~\ref{fig:pairmag} we show the mean absolute residuals between
these primary and secondary magnitudes as a function of magnitude (black dots).
We also plot the Poissonian uncertainties for comparison (solid red line).
We find the mean residual and standard deviation to be $0.03\pm0.04$~mag
across the magnitude ranges 13 to 18 ($r$)
and 12 to 17 ($i$, \ha),
which is consistent with the accuracy of the calibration.
Stars fainter than this range appear to be dominated by photon noise (red line),
while stars at the bright end appear to suffer from large systematic
errors due to saturation effects.

In Fig.~\ref{fig:pairmag_a10} we show 
a similar comparison between the primary and secondary detections,
but this time we have only included sources which are flagged
as \emph{a10point} in the catalogue.
We do not observe an improvement in the average residuals as a function of magnitude,
but the number of outliers has decreased markedly
(evidenced by the shorter error bars which denote the standard deviation of the absolute residuals).
We conclude that the \emph{a10point} quality criteria are effective
at reducing the level of the systematic errors,
while also removing the inaccurate data at the bright and faint end.

\begin{figure}
\captionsetup[subfigure]{labelformat=empty}
\begin{subfigure}[b]{\linewidth}
\centering
\includegraphics[width=0.9\textwidth]{figures/repeatability/repeatability.pdf}
\caption{}
\label{fig:pairmag}
\end{subfigure}
\hspace{0.5cm}
\begin{subfigure}[b]{\linewidth}
\centering
\includegraphics[width=0.9\textwidth]{figures/repeatability/repeatability-a10point.pdf}
\caption{}
\label{fig:pairmag_a10}
\end{subfigure}
\caption{Photometric repeatability as a function of magnitude
         for all sources in the catalogue (panel a)
         and for the \emph{a10point} sources alone (panel b).
         Black dots show the mean absolute residuals
         between the primary and the secondary detections.
         The error-bars show the standard deviations.
         The red trend line shows the average Poissonian uncertainties
         we derived in Fig.~\ref{fig:uncertainties}.
         We find that the \emph{a10point} quality criteria are successful
         at removing objects with large residuals.
    }
\end{figure}
\begin{figure*}
    \vspace{1cm}
    \includegraphics[width=\textwidth]{figures/sourcecount/sourcecount.pdf} 
    \caption{Number of entries in the IPHAS DR2 source catalogue
    as a function of Galactic longitude.
    The upper blue line shows the number of sources
    counted in 1\degr-wide longitude bins.
    The lower red line uses the same binning
    but includes only those sources 
    for which the \emph{deblend} flag is {\sc false}, 
    i.e. unconfused sources for which the CASU pipeline
    did not have to apply a deblending procedure.
    In both cases we counted only those sources
    in the latitude range $-5\degr<b<+5\degr$,
    such that one may obtain a rough guide to source density
    by dividing the counts by 10~deg$^2$.
    The global variations in the source counts
    traces the structure of the Galaxy
    and the distribution of foreground extinction,
    but is also affected by instrumental effects
	such as variations in the observed depth
	and completeness (see text).
   }
    \label{fig:sourcecount}
\end{figure*}

@@

\vspace{0.5cm}

>>>>

An external validation of the astrometry 
has been carried out by comparing our positions against the
United States Naval Observatory CCD Astrograph Catalog 
\citep[UCAC4;][]{UCAC4}.
Fig.~\ref{fig:astrometry} shows the distribution
of the offsets against UCAC4
computed for 1.3 million stars
in the magnitude range $13 < r < 15$,
which is the range where both surveys overlap
and where the formal mean error of UCAC4
is better than 50~mas \citep{UCAC3}.
We find the mean position offset to be $0\farcs 09\pm0\farcs 07$,
which is satisfactory for our purposes.
%We note that both UCAC4 and 2MASS -- and hence also IPHAS --
%not 100% true due to bulk pm of calibrators - Mike
%are calibrated in the International Celestial Reference System (ICRS).
%We note that the mean residual is  
%is $-0.006 \pm 0.086$ in R.A.
%and $0.012\pm0.073$ in Dec.

@@

The entire data set -- comprising 2.5~terabyte of FITS files --
was then transferred to the University of Hertfordshire
for the purpose of transforming the raw
detection tables into a source catalogue which
(i) is quality-controlled,
(ii) is homogeneously calibrated, and 
(iii) contains user-friendly columns and warning flags.
It is these post-processing steps which distinguish this release
from the IDR, which (i) enforced less stringent quality limits,
(ii) did not offer a global calibration,
and (iii) provided a catalogue in which duplicate detections
of unique sources were not flagged.
These improvements are explained next.}

@@

%\newpage

@@

\begin{table*}
\vspace{2.5cm}
\caption{\bf Recommended quality criteria 
for selecting objects with accurate colours.
These criteria serve to identify objects detected 
at S/N$\,>\,$10 in all three bands without being saturated,
and with the added requirement that the photometric measurements
need to be consistent across different aperture sizes.
The 86~million objects which satisfy these criteria
have been flagged in the catalogue using the column named \emph{a10}
(short for ``all-band 10$\sigma$'').}
\begin{tabular}{p{8cm}lp{6cm}}
  \hline
  Quality criterion & Rows passed & Description \\
  \hline
  $rErr\,<\,0.1$ {\sc and} $iErr\,<\,0.1$ {\sc and} $haErr\,<\,0.1$ &
  109 million (50\%) &
  Require the photon noise to be less than 
  0.1 mag in all bands (i.e. S/N$>$10).
  This implicitly requires a detection in all three bands.  \\
  {\sc not} \emph{saturated} &
  158 million (72\%) &
  The brightness must not exceed the nominal saturation limits.
  \\
  $|r- rAperMag1| 
  < 3\sqrt{rErr^2 + rAperMag1Err^2} 
  + 0.03$ &
  176 million (80\%) &
  Require the $r$ magnitude measured 
  in the default 2\farcs 3-diameter aperture
  to be consistent with the measurement 
  made in the smaller 1\farcs 2 aperture,
  albeit tolerating a 0.03 mag systematic error.
  This will reject sources for which the background
  subtraction or the deblending procedure
  was not performed reliably. \\
  $|i- iAperMag1| 
  < 3\sqrt{iErr^2 + iAperMag1Err^2} 
  + 0.03$ &
  183 million (84\%) &
  Same as above for $i$. \\
  $|ha- haAperMag1| < 
  3\sqrt{haErr^2 + haAperMag1Err^2} 
  + 0.03$ &
  158 million (72\%) &
  Same as above for \ha. \\
  \hline
  All of the above (flagged as {\bf\emph{a10}}) &
  86 million (39\%) & \\
  \hline
\end{tabular}
\label{tab:a10}
\vspace{2.5cm}
\caption{\bf Additional quality criteria which are recommended
for applications which require objects to be
single, unconfused point sources with accurate colours.
The 59 million sources which satisfy these criteria 
have been flagged using the column named \emph{a10point}.}
\begin{tabular}{p{8cm}lp{6cm}}
  \hline
  Quality criterion & Rows passed & Description \\
  \hline
   \emph{a10} &
   86 million (39\%) &
   The object must satisfy the criteria for accurate colours listed in Table~\ref{tab:a10}. \\

@@

@@

To appraise the level at which our photometry is affected by such systematics,
we can exploit the secondary measurements which are present in the catalogue
(i.e. \emph{r} vs \emph{r2}, \emph{i} vs \emph{i2}, \emph{ha} vs \emph{ha2}).
In Fig.~\ref{fig:pairmag}a we show the mean absolute residuals between
these primary and secondary magnitudes as a function of magnitude (black dots).
We also plot the Poissonian uncertainties for comparison (solid red line).
We find the mean residual and standard deviation to be $0.03\pm0.04$~mag
across the magnitude ranges 13 to 18 ($r$)
and 12 to 17 ($i$, \ha),
which is consistent with the accuracy of the calibration.
Stars fainter than this range appear to be dominated by photon noise (red line),
while stars at the bright end appear to suffer from large systematic
errors due to saturation effects.

\begin{figure*}
	\vspace{1cm}
    \begin{minipage}[b]{\linewidth}
        \includegraphics[width=0.5\textwidth]{figures/repeatability/repeatability.pdf} 
        \includegraphics[width=0.5\textwidth]{figures/repeatability/repeatability-a10point.pdf}
    \end{minipage}
    \caption{Photometric repeatability as a function of magnitude
         for all sources in the catalogue (panel a)
         and for the \emph{a10point} sources alone (panel b).
         Black dots show the mean absolute residuals
         between the primary and the secondary detections.
         The error-bars show the standard deviations.
         The red trend line shows the average Poissonian uncertainties
         we derived in Fig.~\ref{fig:uncertainties}.
         We find that the \emph{a10point} quality criteria are successful
         at removing objects with large residuals.}
    \label{fig:pairmag}

    \vspace{2cm}
    \includegraphics[width=\textwidth]{figures/sourcecount/sourcecount.pdf} 
    \caption{Number of entries in the IPHAS DR2 source catalogue
    as a function of Galactic longitude.
    The upper blue line shows the number of sources
    counted in 1\degr-wide longitude bins.
    The lower red line uses the same binning
    but includes only those sources 
    for which the \emph{deblend} flag is {\sc false}, 
    i.e. unconfused sources for which the CASU pipeline
    did not have to apply a deblending procedure.
    In both cases we counted only those sources
    in the latitude range $-5\degr<b<+5\degr$,
    such that one may obtain a rough guide to source density
    by dividing the counts by 10~deg$^2$.
    The global variations in the source counts
    traces the structure of the Galaxy
    and the distribution of foreground extinction,
    but is also affected by instrumental effects
	such as variations in the observed depth
	and completeness (see text).
   }
    \label{fig:sourcecount}
\end{figure*}

In Fig.~\ref{fig:pairmag}b we show 
a similar comparison between the primary and secondary detections,
but this time we have only included sources which are flagged
as \emph{a10point} in the catalogue.
We do not observe an improvement in the average residuals as a function of magnitude,
but the number of outliers has decreased markedly
(evidenced by the shorter error bars which denote the standard deviation of the absolute residuals).
We conclude that the \emph{a10point} quality criteria are effective
at reducing the level of the systematic errors,
while also removing the inaccurate data at the bright and faint end.

@@

%\vspace{0.5cm}
